---
title: "Credit Risk ML: Model Performance Analysis"
subtitle: "ILAB OeNB — VAE-Augmented Default Prediction"
author: "Anastasiia Pryshchepa"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: false
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(knitr)
library(kableExtra)
```


# Executive Summary

This document explains why certain model configurations outperform others in predicting corporate defaults using Austrian balance sheet data. The analysis covers 30 model variants across 5 feature sets and 3 algorithms.

**Key Finding:** Adding time dynamics features provides the largest performance gain (+8 AUC points). VAE latent features add minimal incremental value. GLM shows the best temporal stability despite lower peak performance.

# The Model Comparison Framework

## Feature Sets Tested

We tested five progressively complex feature configurations:

| Group | Feature Set | Description |
|:------|:------------|:------------|
| 01 | Raw Balance Sheet | 11 balance sheet items + sector information |
| 02 | Financial Ratios | 20 computed ratios + sector-relative features |
| 03 | Ratios + Time Dynamics | Ratios + exponential moving averages, trends, momentum |
| 04 | Ratios + TD + VAE | Group 03 features + 32 VAE latent dimensions |
| 05 | VAE Only | 32 VAE latent features + categorical sector |

## Algorithms Compared

- **XGBoost Manual**: Gradient boosted trees with hyperparameter tuning
- **AutoGluon**: Automated ML with model ensembling
- **GLM**: Elastic net logistic regression (interpretable baseline)

## Evaluation Splits

- **Out-of-Sample (OoS)**: Random 80/20 split, ~70k test observations
- **Out-of-Time (OoT)**: Temporal split, ~208k test observations from later period

# Why Time Dynamics Features Matter Most

## The Performance Hierarchy

```{r performance-table}
perf_data <- data.frame(
  `Feature Set` = c("Ratios + Time Dynamics", "Ratios + TD + VAE", 
                    "VAE Latent Only", "Financial Ratios", "Raw Balance Sheet"),
  `OoS AUC` = c("0.914", "0.911", "0.847", "0.830", "0.817"),
  `OoT AUC` = c("0.824", "0.825", "0.777", "0.753", "0.692"),
  `Gain vs Raw` = c("+9.7 pts", "+9.4 pts", "+3.0 pts", "+1.3 pts", "baseline")
)

kable(perf_data, booktabs = TRUE, col.names = c("Feature Set", "OoS AUC", "OoT AUC", "Gain vs Raw"),
      caption = "Mean AUC by Feature Set") %>%
  kable_styling(latex_options = c("striped", "hold_position")) %>%
  row_spec(1, bold = TRUE)
```

The table reveals a clear pattern:

1. **Time dynamics adds ~8 AUC points** over static ratios (Group 03 vs Group 02)
2. **VAE adds virtually nothing** over time dynamics (Group 04 $\approx$ Group 03)
3. **VAE alone matches ratios** — the autoencoder compresses financial structure effectively

## Why history_length Dominates

The single most important feature is `history_length` — the number of consecutive years a firm appears in the dataset. Its importance scores range from 0.19 to 0.26 across models.

This feature captures multiple risk signals:

- **Young firms** have short histories and higher default rates
- **Disappearing firms** often exit the dataset shortly before default
- **Stable firms** with long histories demonstrate survival capability

The feature may also capture data artifacts — firms entering the dataset just before default will have short observable histories by construction.

## Time-Based Features Capture Deterioration

Beyond history length, the time dynamics features add:

- **Exponential moving averages** (`expmean_r*`): Weighted recent performance
- **Momentum indicators** (`momentum_r*`): Rate of change in ratios
- **Peak-drop metrics** (`peak_drop_r*`): Distance from historical best
- **Temporal position** (`time_index`): Macro cycle context

These features allow models to distinguish between a firm that has always been weak versus one that is actively deteriorating — a critical distinction for default prediction.

# Why VAE Features Add Limited Value

## The VAE Architecture

The Variational Autoencoder was trained to compress 20 financial ratios into 32 latent dimensions. The hypothesis was that these latent features might capture non-linear financial structures invisible to standard ratios.

## Why It Didn't Help

Three factors explain the VAE's limited contribution:

1. **Posterior collapse**: 13 of 32 latent dimensions collapsed (zero variance), meaning the autoencoder learned to ignore nearly half its capacity

2. **Redundancy with sector features**: The VAE latent dimensions appear to encode sector-relative financial health — information already captured by `secRank_*` features in the ratio set

3. **Already saturated signal**: With time dynamics already achieving 0.91 AUC, there may be limited room for improvement without fundamentally new information (e.g., payment behavior, supplier data)

## When VAE-Only Works Reasonably Well

Interestingly, the VAE-only models (Group 05) achieve 0.847 AUC — competitive with the financial ratios baseline. This suggests the autoencoder successfully compresses financial structure, but doesn't add information beyond what explicit ratios provide.

# Algorithm Comparison: The Stability-Performance Trade-off

## Performance Summary

```{r algo-table}
algo_data <- data.frame(
  Algorithm = c("XGBoost", "AutoGluon", "GLM"),
  `Best OoS` = c("0.920", "0.916", "0.905"),
  `Best OoT` = c("0.811", "0.810", "0.859"),
  `OoS-OoT Gap` = c("10.9 pts", "10.6 pts", "4.6 pts"),
  Interpretation = c("Highest peak, largest drift", "Strong ensemble, moderate drift", "Lower peak, most stable")
)

kable(algo_data, booktabs = TRUE, 
      col.names = c("Algorithm", "Best OoS AUC", "Best OoT AUC", "OoS→OoT Gap", "Interpretation"),
      caption = "Algorithm Performance Comparison") %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

## The Concept Drift Problem

All models show substantial OoS-to-OoT degradation, but GLM degrades least:

- **XGBoost/AutoGluon** overfit to the training distribution. Their flexible function approximators capture patterns that don't generalize temporally.

- **GLM** constrains model complexity through regularization. While this limits peak performance, it produces more robust predictions across time periods.

For production credit risk models where temporal stability matters, GLM may be preferable despite its lower OoS ceiling.

# Confusion Matrix Insights

## The Class Imbalance Challenge

With only 0.83% default rate, achieving high recall requires accepting low precision:

```{r cm-table}
cm_data <- data.frame(
  Model = c("04a_XGBoost", "03a_XGBoost", "03a_AutoGluon", "03a_GLM"),
  Recall = c("88.2%", "87.4%", "85.1%", "81.5%"),
  Precision = c("4.1%", "4.0%", "4.4%", "4.6%"),
  F1 = c("7.8%", "7.7%", "8.4%", "8.7%"),
  FP = c("11,942", "12,079", "10,721", "9,801")
)

kable(cm_data, booktabs = TRUE,
      caption = "Confusion Matrix Metrics at Optimal Threshold") %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

At the Youden-optimal threshold:

- Models catch 81-88% of actual defaults (high recall)
- Only 4% of flagged firms actually default (low precision)
- For every true default caught, 20-25 false alarms occur

This trade-off is inherent to rare-event prediction. In practice, banks may prefer higher thresholds (fewer false positives) or implement tiered review processes.

# Recommendations

Based on this analysis, we recommend:

1. **Use Group 03 (Ratios + Time Dynamics)** as the primary feature set. VAE adds complexity without meaningful benefit.

2. **Prefer GLM for production** if temporal stability matters more than peak OoS performance. The 5-point OoS-OoT gap vs 11-point gap is significant.

3. **Investigate `history_length`** further — determine whether its predictive power reflects genuine risk signal or data artifact.

4. **Recalibrate OoT probabilities** — negative Brier Skill Scores indicate miscalibration that isotonic regression could address.

5. **Add macroeconomic covariates** (interest rates, GDP growth, unemployment) to reduce concept drift effects.

# Appendix: Key Metrics Definitions

- **AUC-ROC**: Area under the Receiver Operating Characteristic curve. Measures discrimination regardless of threshold.

- **Gini**: 2$\times$AUC - 1. Ranges from 0 (random) to 1 (perfect).

- **Recall@FPR**: Fraction of defaults captured at a given false positive rate. More operationally relevant than AUC.

- **Brier Score**: Mean squared error of probability predictions. Lower is better.

- **BSS (Brier Skill Score)**: Improvement over climatology baseline. Negative values indicate worse-than-naive calibration.

- **ECE (Expected Calibration Error)**: Average gap between predicted probabilities and observed frequencies across decile bins.

